---
output: github_document
---

# Pet Adoption

#### By Amelia Humphrey & Mariana Correa

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Description:

Pet Adoption: <https://www.kaggle.com/datasets/rabieelkharoua/predict-pet-adoption-status-dataset>

```{r}
library(readr)
pets <- read_csv("pet_adoption_data.csv")
head(pets)
summary(pets)
dim(pets)

```

The dataset we chose provides a comprehensive look into various factors that can influence the likelihood of a pet being adopted from a shelter, covering various characteristics and attributes.

We do foresee a lot of data cleaning. We will be checking for missing values, duplicates, and outliers that could skew the data. We also plan to change columns with binary '1' and '0' values to boolean values.

```{r}
# Load libraries
library(dplyr)
library(ggplot2)
library(tidyverse)

# Check for missing values
colSums(is.na(pets))
```
There are no missing values in any of the columns of this dataset.

```{r}
# Count duplicates
sum(duplicated(pets))
```
There are no duplicate rows in this dataset.

```{r}
# Convert binary
pets$Vaccinated <- as.logical(pets$Vaccinated)
pets$HealthCondition <- as.logical(pets$HealthCondition)
pets$PreviousOwner <- as.logical(pets$PreviousOwner)
pets$AdoptionLikelihood <- as.logical(pets$AdoptionLikelihood)
```
The columns `Vaccinated`, `HealthCondition`, `PreviousOwner` and `AdoptionLikelihood` were changed from "0" and "1" to boolean values.

```{r}
# Check datset
str(pets)
summary(pets)
```

Some summaries we plan to explore are the averages of numerical values between the different types of pet, and among different breeds. We also plan to investigate the data by exploring the different pet type and breeds, as well as the distribution of different color, weight, and type.


```{r}

#different types of pet 
unique(pets$PetType)

pets |> 
  filter(PetType == "Dog") |> 
  distinct(Breed)
pets |> 
  filter(PetType == "Cat") |> 
  distinct(Breed)
pets |> 
  filter(PetType == "Bird") |> 
  distinct(Breed)
pets |> 
  filter(PetType == "Rabbit") |> 
  distinct(Breed)

pets |>
  ggplot(aes(x = PetType, fill = PetType)) +
  geom_bar() +
  ggtitle("Number of Pets by Pet Type") +
  xlab("Pet Type") 

  
  


```

### Research Questions:

1.  What factors impact the likelihood of adoption? Does a longer time in shelter decrease the likelihood of adoption? What type of animal is most likely to get adopted? How does the likelihood vary between breed for the same pet type? Is there a certain pet color that is more likely to be adopted?

```{r}
#On average how likely is a pet to be adopted?
mean(pets$AdoptionLikelihood)


#The time in shleter is not related to the likelihood of adoption
cor(pets$TimeInShelterDays, pets$AdoptionLikelihood)

#which pet is most likely to be adopted?

 pets |> 
  group_by(PetType) |> 
  summarize(
    n = n(),
    adoptionRate = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(adoptionRate))
 
 
RateAdoption <- pets |> 
  group_by(PetType) |> 
  summarize(
    adoptionRate = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(adoptionRate))

#SOOOO UGLY FIXXXX
RateAdoption |> 
  ggplot(aes(x=adoptionRate, y = PetType)) +
  geom_col(fill = "#EE1289") 



#Does breed play a role 
#Note Rabbit and Birds were not considered as they each only have one Breed
pets |> 
  filter(PetType == "Dog") |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    rate  = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate)) |> 
  ggplot(aes(x = Breed, y = rate)) + 
  geom_col(fill = "darkslategray2") + 
  ggtitle("Breed and Adoption Rate for Dogs")+
  ylab("Adoption Rate")

pets |> 
  filter(PetType == "Cat") |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    rate  = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate))  |> 
  ggplot(aes(x = Breed, y = rate)) + 
  geom_col(fill = "darkslategray") + 
  ggtitle("Breed and Adoption Rate for Cats")+
  ylab("Adoption Rate")


#Does color

pets |> 
  group_by(Color) |> 
  summarise(
    n = n(),
    rate = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate)) |> 
  ggplot(aes(x = rate, y= Color)) + 
  geom_col(fill = "dodgerblue3") + 
  ggtitle("Color and adoption")


#facet moment
pets |> 
  group_by(PetType, Color) |> 
  summarise(
    n = n(),
    rate = mean(AdoptionLikelihood)
  ) |> 
  ggplot(aes(x= Color, y = rate, fill = Color))  +
  geom_col() + 
  scale_fill_manual(values= c("Black" = "black", "Brown" = "burlywood4", "Gray" = "gray80", "Orange" = "darkorange" , "White" = "cornsilk" )) +
  facet_wrap(~PetType) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  ggtitle("Color vs Average Adoption Likelihood")+
  ylab("Average Adoption Likelihood")


```

ADD IN A CONCLUSION BASED ON THE DATA and the graphs 

Note: the average was taken as each PetType does not have the same amount of entries. Taking the mean make the data more meaningful in a regardless of the amount of data for each type of animal. 


2.  What factor has the largest impact on time spent in the shelter? Is age a contributor? What breed is more likely to spend a long time in the shelter? Does health condition impact the time spent in shelter?


```{r}

#Putting more emphasis on the median is important as there are many outlier that can skew the mean results 

pets |> 
  ggplot(aes(x = TimeInShelterDays))  +
  geom_histogram(binwidth = 3, fill = "darkslategray4") + 
  ggtitle("Time Spent in Shelter") + 
  ylab("Number of Pets")


pets |> 
  ggplot(aes(x = TimeInShelterDays))  +
  geom_histogram(binwidth = 3) + 
  facet_wrap(~PetType) +
  ggtitle("Time Spent in Shelter") + 
  ylab("Number of Pets")  

pets |> 
  group_by(PetType) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) |> 
  arrange(desc(medianDays)) |> 
  ggplot(aes(x = PetType, y = medianDays)) + 
  geom_col(fill = "darkorchid4") + 
  ggtitle("Median Time in Shelter by Pet Type") + 
  ylab("Median") 



pets |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) |> 
  arrange(desc(medianDays)) |> 
  ggplot(aes(x = Breed, y = medianDays)) + 
  geom_col(fill = "darkolivegreen") + 
  ggtitle("Median Time in Shelter by Breed") + 
  ylab("Median") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

#making a facet
cor(pets$AgeMonths, pets$TimeInShelterDays, use = "complete.obs")
cor(pets$WeightKg, pets$TimeInShelterDays, use = "complete.obs")
cor(pets$AdoptionFee, pets$TimeInShelterDays, use = "complete.obs")

#shows the randomness/lack of correlation
pets |> 
  ggplot(aes(x = AgeMonths, y = TimeInShelterDays)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "dodgerblue3") +
  facet_wrap(~PetType) + 
  ggtitle("Age vs Time in Shelter") + 
  xlab("Age (months)")  +
  ylab("Time in Shelter (days)")

#Does being old increase shelter time 
#over 7 can be considered old for a pet

#add color
#add axises etitles and title
pets |> 
  mutate(AgeGroups = if_else(AgeMonths <24, "Young(<2 years old)", if_else(AgeMonths < 84, "Adult (2-7 years)", "Old(>7 years old"))) |> 
  group_by(AgeGroups) |> 
  summarise(
    median = median(TimeInShelterDays)
  )  |> 
  ggplot(aes(x = AgeGroups, y = median)) + 
  geom_col()


#long stay calculate top 10% of the the total time shelter

pets2 <- pets |> 
  mutate(LongStay = TimeInShelterDays >= quantile(TimeInShelterDays, 0.9, na.rm = TRUE))


#Is there a breed that typical stays longer 

LongStayBreed <- pets2 |>
  group_by(Breed) |> 
  summarise(
    n = n(),
    medianDays = median(TimeInShelterDays), longStayRate = mean(LongStay),
    avgDays  =  mean(TimeInShelterDays)
  ) |> 
  arrange(desc(medianDays))
LongStayBreed

#why n 30
#mess with the prettiness of the graph
LongStayBreed |> 
  filter(n >=30) |> 
  arrange(desc(longStayRate)) |> 
  ggplot(aes(x = longStayRate, y = reorder(Breed, longStayRate))) +
  geom_bar(stat = "identity") + 
  coord_flip()
  

#Does having a healthCondition impact shelter time?

pets |> 
  group_by(HealthCondition) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) 


  

#graph
pets |> 
  ggplot(aes(x = HealthCondition, y = TimeInShelterDays)) +
  geom_boxplot() + 
  xlab("Health Condition")

#add titles and colors 
pets |> 
  ggplot(aes(x = HealthCondition, y = TimeInShelterDays, fill = HealthCondition)) +
  geom_boxplot() +
  coord_flip()+
  facet_wrap(~PetType)  +
  xlab("Health Condition")

```






3.  Is there a breed that has a larger adoption fee? Does the adoption fee have any correlation to the size of the pet? Do pets with higher adoption fees have better health conditions/are vaccinated? Does the color of the pet impact the adoption fee?

``` {r}

# Is there a breed that has a larger adoption fee?
# Using median fee to avoid skewness from outliers
pets |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    avgFee = mean(AdoptionFee),
    medianFee = median(AdoptionFee)
  ) |> 
  arrange(desc(medianFee)) |> 
  ggplot(aes(x = reorder(Breed, medianFee), y = medianFee)) + 
  geom_col(fill = "lightseagreen") + 
  coord_flip() +
  ggtitle("Median Adoption Fee by Breed") +
  xlab("Breed") +
  ylab("Median Adoption Fee ($)")

# Does the adoption fee have any correlation to the size of the pet?
cor(pets$WeightKg, pets$AdoptionFee, use = "complete.obs")
pets |> 
  ggplot(aes(x = WeightKg, y = AdoptionFee)) +
  geom_point(alpha = 0.5, color = "sienna") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  facet_wrap(~PetType) + 
  ggtitle("Adoption Fee vs. Pet Weight by Pet Type") +
  xlab("Weight (Kg)") +
  ylab("Adoption Fee ($)")

# Do pets with higher adoption fees have better health conditions/are vaccinated?
pets |> 
  ggplot(aes(x = HealthCondition, y = AdoptionFee, fill = HealthCondition)) +
  geom_boxplot() +
  ggtitle("Adoption Fee by Health Condition") +
  xlab("Health Condition") +
  ylab("Adoption Fee ($)")

pets |> 
  ggplot(aes(x = Vaccinated, y = AdoptionFee, fill = Vaccinated)) +
  geom_boxplot() +
  scale_fill_manual(values = c("FALSE" = "tomato", "TRUE" = "springgreen4")) +
  ggtitle("Adoption Fee by Vaccination Status") +
  xlab("Vaccinated") +
  ylab("Adoption Fee ($)")

# Does the color of the pet impact the adoption fee?
pets |> 
  ggplot(aes(x = reorder(Color, AdoptionFee, FUN = median), y = AdoptionFee, fill = Color)) +
  geom_boxplot() +
  scale_fill_manual(values= c("Black" = "grey30", "Brown" = "burlywood4", 
                              "Gray" = "gray70", "Orange" = "darkorange", "White" = "wheat1")) +
  ggtitle("Adoption Fee Distribution by Color") +
  xlab("Color") +
  ylab("Adoption Fee ($)") +
  theme(legend.position = "none")
```

4.  Which type of pets are most likely to be healthy (HealthCondition = 1), and how does that connect to the time spent in the shelter? Are medical consitions more common among older animals across all pet types? Are certain breeds more likely to have a medical condition?

``` {r}
# Which type of pets are most likely to be healthy (HealthCondition = FALSE)?
pets |> 
  group_by(PetType) |> 
  summarise(
    n = n(),
    # Since TRUE = Medical Condition, the proportion of healthy pets is 1 minus the mean
    HealthRate = 1 - mean(HealthCondition) 
  ) |> 
  arrange(desc(HealthRate)) |> 
  ggplot(aes(x = reorder(PetType, HealthRate), y = HealthRate)) +
  geom_col(fill = "mediumpurple3") +
  ggtitle("Proportion of Healthy Pets by Pet Type") +
  xlab("Pet Type") +
  ylab("Proportion Healthy (HealthCondition = FALSE)")

# How does that connect to the time spent in the shelter?
pets |> 
  ggplot(aes(x = HealthCondition, y = TimeInShelterDays, fill = HealthCondition)) +
  geom_boxplot() +
  scale_fill_manual(values = c("FALSE" = "springgreen4", "TRUE" = "tomato")) +
  facet_wrap(~PetType) +
  ggtitle("Time in Shelter by Medical Condition and Pet Type") +
  xlab("Has Medical Condition (TRUE = Yes, FALSE = No)") +
  ylab("Time in Shelter (Days)")

# Are medical conditions more common among older animals across all pet types?
pets |> 
  mutate(AgeGroups = if_else(AgeMonths < 24, "Young(<2 years)", 
                     if_else(AgeMonths < 84, "Adult(2-7 years)", "Old(>7 years)"))) |> 
  mutate(AgeGroups = factor(AgeGroups, levels = c("Young(<2 years)", "Adult(2-7 years)", "Old(>7 years)"))) |>
  group_by(AgeGroups, PetType) |> 
  summarise(
    n = n(),
    # Since TRUE = Medical Condition, the mean gives the direct proportion of sick pets
    MedicalConditionRate = mean(HealthCondition), 
    .groups = 'drop'
  ) |> 
  ggplot(aes(x = AgeGroups, y = MedicalConditionRate, fill = PetType)) +
  geom_col(position = "dodge") + 
  ggtitle("Rate of Medical Conditions by Age Group and Pet Type") +
  xlab("Age Group") +
  ylab("Proportion with Medical Condition")

# Are certain breeds more likely to have a medical condition?
pets |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    MedicalConditionRate = mean(HealthCondition)
  ) |> 
  arrange(desc(MedicalConditionRate)) |> 
  ggplot(aes(x = reorder(Breed, MedicalConditionRate), y = MedicalConditionRate)) +
  geom_col(fill = "indianred") +
  coord_flip() +
  ggtitle("Proportion of Pets with Medical Conditions by Breed") +
  xlab("Breed") +
  ylab("Proportion with Medical Condition")
```

To investigate these questions, we plan on looking at boxplots, histograms and scatter plots and perform linear regressions to identify trends and correlations within the data.

