---
output: github_document
---

# Pet Adoption

#### By Amelia Humphrey & Mariana Correa

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(dplyr)
library(ggplot2)
library(tidyverse)
library(readr)
```

## Introduction

The goal of our project is to explore the 'Pet Adoption' dataset to better understand which pets have the best likelihood of adoption. Animal shelters across the world are overcrowded, making it important to identify what influences adoption. Our aim is to find factors that impact a pet's adoption likelihood, so shelters can better represent underadopted pets. This analyze could help bring awareness to pet adoption and increase the adoption rates for underadopted pets.

To achieve our goal, we are exploring the following key questions:

-   What factors affect a pet's likelihood of adoption?

-   Which factor has the strongest influence on the length of time a pet spends in the shelter?

-   How do adoption fees relate to a pet's likelihood of adoption?

-   How does a pet's health condition relate to other factors in pet adoption?

We hope these findings will help us to conclude how shelters can best help pets to b e adopted.

### Data Description

We are using a kaggle dataset created by Rabie El kharoua in 2024. This dataset provides a comprehensive look into various factors that can influence the likelihood of a pet being adopted from a shelter, covering various characteristics and attributes.

Pet Adoption Link: <https://www.kaggle.com/datasets/rabieelkharoua/predict-pet-adoption-status-dataset>

Citation:

Rabie El kharoua. (2024). 🐾 Predict Pet Adoption Status Dataset 🐾 [Data set]. Kaggle. <https://doi.org/10.34740/KAGGLE/DS/5242440>

```{r}

pets <- read_csv("pet_adoption_data.csv")
head(pets)
summary(pets)
dim(pets)

```

## Cleaning

To clean the dataset, we checked for missing values, duplicates, and outliers that could skew the data. We changed columns with binary '1' and '0' values to boolean values.

```{r}
# Check for missing values
colSums(is.na(pets))
```

There are no missing values in any of the columns of this dataset.

```{r}
# Count duplicates
sum(duplicated(pets))
```

There are no duplicate rows in this dataset.

```{r}
# Convert binary
pets$Vaccinated <- as.logical(pets$Vaccinated)
pets$HealthCondition <- as.logical(pets$HealthCondition)
pets$PreviousOwner <- as.logical(pets$PreviousOwner)
pets$AdoptionLikelihood <- as.logical(pets$AdoptionLikelihood)
```

The columns `Vaccinated`, `HealthCondition`, `PreviousOwner` and `AdoptionLikelihood` were changed from "0" and "1" to boolean values.

```{r}
# Check datset
head(pets)
summary(pets)
dim(pets)
```

Our variable after cleaning:

1.  **PetID:** Unique identifier for each pet.

2.  **PetType:** Type of pet (e.g., Dog, Cat, Bird, Rabbit).

3.  **Breed:** Specific breed of the pet.

4.  **AgeMonths:** Age of the pet in months.

5.  **Color:** Color of the pet.

6.  **Size:** Size category of the pet (Small, Medium, Large).

7.  **WeightKg:** Weight of the pet in kilograms.

8.  **Vaccinated:** Vaccination status of the pet (0 - Not vaccinated, 1 - Vaccinated).

9.  **HealthCondition:** Health condition of the pet (0 - Healthy, 1 - Medical condition).

10. **TimeInShelterDays:** Duration the pet has been in the shelter (days).

11. **AdoptionFee:** Adoption fee charged for the pet (in dollars).

12. **PreviousOwner:** Whether the pet had a previous owner (0 - No, 1 - Yes).

13. **AdoptionLikelihood:** Likelihood of the pet being adopted (0 - Unlikely, 1 - Likely).

An initial view of the number of pets in the data type and what type of animals are in the animal shelter.

```{r}

pets |>
  ggplot(aes(x = PetType, fill = PetType)) +
  geom_bar() +
  ggtitle("Number of Pets by Pet Type") +
  xlab("Pet Type") 

pets |>
  ggplot(aes(x = Breed, fill = Breed)) +
  geom_bar() +
  ggtitle("Number of Pets by Pet Type") +
  xlab("Breeds") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

These graphs show the pet type and breed distribution in the dataset. The most common breeds in the datasets are parakeet and rabbit. The pet type that is most common at the shelter is dogs followed closest by cats.

### Research Questions:

1.  **What factors impact the likelihood of adoption? Does a longer time in shelter decrease the likelihood of adoption? What type of animal is most likely to get adopted? How does the likelihood vary between breed for the same pet type? Is there a certain pet color that is more likely to be adopted?**

```{r}
#On average how likely is a pet to be adopted?
mean(pets$AdoptionLikelihood)


#is shelter time correlated to adoptionLikelihood?
cor(pets$TimeInShelterDays, pets$AdoptionLikelihood)

#which pet type is most likely to be adopted?

 RateAdoption <-pets |> 
  group_by(PetType) |> 
  summarize(
    n = n(),
    adoptionRate = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(adoptionRate))
 RateAdoption
 
 RateAdoption |> 
  ggplot(aes(x=adoptionRate, y = PetType)) +
  geom_col(fill = rgb(17, 202, 160, maxColorValue = 255)) + 
  ggtitle("Pet Type and Adoption Rate") +
  xlab("Adoption Rate")  +
  ylab("Pet Type")
```

A pet is on average, 32.8% likely to be adopted. The time in the shelter's correlation to adoption likelihood is practically nonexistent with a value of 0.009. The type of pet most likely to be adopted is dogs, at 46.4%, and the least likely pet to be adopted is rabbits, at 25.4%.

```{r}
pets |> 
  filter(PetType == "Dog") |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    rate  = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate)) |> 
  ggplot(aes(x = Breed, y = rate)) + 
  geom_col(fill = rgb(196, 146, 177, maxColorValue = 255)) + 
  ggtitle("Breed and Adoption Rate for Dogs")+
  ylab("Adoption Rate")  +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

pets |> 
  filter(PetType == "Cat") |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    rate  = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate))  |> 
  ggplot(aes(x = Breed, y = rate)) + 
  geom_col(fill = rgb(17, 202, 160, maxColorValue = 255)) + 
  ggtitle("Breed and Adoption Rate for Cats")+
  ylab("Adoption Rate")

```

These graphs show the adoption likelihoods for different types of breeds by pet type. Birds and rabbits do not have graphs, as each only have one breed. From the dog graph, it is clear that Labradors are the most likely dog breed to be adopted with around 73% likely. The cat graph displays that Siamese are more likely to be adopted in comparison to Persian cats. These graphs show that Persian cats and Golden Retrievers could use more marketing to help increase their adoption numbers.

```{r}

pets |> 
  group_by(Color) |> 
  summarise(
    n = n(),
    rate = mean(AdoptionLikelihood)
  ) |> 
  arrange(desc(rate)) |> 
  ggplot(aes(x = rate, y= Color, fill = Color)) + 
  geom_col() +
   scale_fill_manual(values= c("Black" = "black", "Brown" = "burlywood4", "Gray" = "gray80", "Orange" = "darkorange" , "White" = "cornsilk" ))+ 
  theme(legend.position = "none") +
  ggtitle("Color and Adoption") + 
  xlab("Adoption Rate")

pets |> 
  filter(PetType == "Dog", Color == "Brown") |> 
  summarize(
    rate = mean(AdoptionLikelihood)
  )


#facet moment
pets |> 
  group_by(PetType, Color) |> 
  summarise(
    n = n(),
    rate = mean(AdoptionLikelihood)
  ) |> 
  ggplot(aes(x= Color, y = rate, fill = Color))  +
  geom_col() + 
  scale_fill_manual(values= c("Black" = "black", "Brown" = "burlywood4", "Gray" = "gray80", "Orange" = "darkorange" , "White" = "cornsilk" )) +
  facet_wrap(~PetType) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  ggtitle("Color vs Average Adoption Likelihood")+
  ylab("Average Adoption Likelihood")

pets |> 
  filter(PetType == "Dog", Color == "Brown") |> 
  summarize(
    rate = mean(AdoptionLikelihood)
  )
```

The top graph shows a general color breakdown by adoption likelihood with orange pets being the most likely and white pets being the least likely. The second graph facets by pet type, showing that brown dogs are the most likely to be adopted. White rabbits are less likely to be adopted than other pets and, within rabbits, less likely than other colors. These findings suggest the white rabbit should be shown first to potential owners in hopes are increased odds of adoption.

```{r}
most_adopted <- pets |> 
  group_by(PetType, Breed, Color) |> 
  summarise(
    rate = mean(AdoptionLikelihood, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  arrange(desc(rate))
most_adopted

#top 15 to plot
most_adopted |> 
  slice_max(rate, n = 15) |> 
  ggplot(aes(x = reorder(paste(PetType, Breed, Color, sep = "/"), rate), y= rate, fill = PetType)) + 
  geom_col()+
  coord_flip()+
  labs(
    title = "Most Adoptable Pet",
    x = "Pet",
    y = "Average Adoption Likilihood"
  ) + theme_minimal()
  
```

This graph displays the top 15 combinations of pet type, breed, and color that are most likely to be adopted. From the graph, it can be extracted that the most likely pet to be adopted in the shelter is a Brown Labrador with over 80%. The least likely dog to be adopted is a white poodle.

2.  **What factor has the largest impact on time spent in the shelter? Is age a contributor? What breed is more likely to spend a long time in the shelter? Does health condition impact the time spent in shelter?**

```{r}
pets |> 
  ggplot(aes(x = TimeInShelterDays))  +
  geom_histogram(binwidth = 6, fill = rgb(196, 146, 177, maxColorValue = 255)) + 
  ggtitle("Time Spent in Shelter") + 
  ylab("Number of Pets")

```

This graph displays how many pets stay for a certain number of days. The graph looks bimodal, with peaks around 15-25 days and 60-65 days. The large drops indicate that adoption has increased.

```{r}
pets |> 
  ggplot(aes(x = TimeInShelterDays))  +
  geom_histogram(binwidth = 6, fill = rgb(196, 146, 177, maxColorValue = 255)) + 
  facet_wrap(~PetType) +
  ggtitle("Time Spent in Shelter") + 
  ylab("Number of Pets")  
```

This graph shows the number of days a pet has stayed faceted by pet type.

```{r}
pets |> 
  group_by(PetType) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) |> 
  arrange(desc(avgDays)) |> 
  ggplot(aes(x = PetType, y = avgDays)) + 
  geom_col(fill = rgb(180, 205, 205, maxColorValue = 255)) + 
  ggtitle("Mean Time in Shelter by Pet Type") + 
  ylab("Mean") 
```

The graph shows the average time a pet type spends in the shelter. Rabbits have the longest average stay time, but all the animals spend about the same amount on average in the shelter (within 10 days).

```{r}
pets |> 
  group_by(Breed) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) |> 
  arrange(desc(avgDays)) |> 
  ggplot(aes(x = Breed, y = avgDays)) + 
  geom_col(fill = rgb(0, 80, 136, maxColorValue = 255)) + 
  ggtitle("Mean Time in Shelter by Breed") + 
  ylab("Mean") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

The graph was created to see if there is a certain breed that has abnormally long stays. The graph shows that all of the breeds are within 10 days of average shelter time. This means that not one type of breed is particularly, not adopted quickly.

```{r}

cor(pets$AgeMonths, pets$TimeInShelterDays, use = "complete.obs")
cor(pets$WeightKg, pets$TimeInShelterDays, use = "complete.obs")
cor(pets$AdoptionFee, pets$TimeInShelterDays, use = "complete.obs")

#shows the randomness/lack of correlation
pets |> 
  ggplot(aes(x = AgeMonths, y = TimeInShelterDays)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "dodgerblue3") +
  facet_wrap(~PetType) + 
  ggtitle("Age vs Time in Shelter") + 
  xlab("Age (months)")  +
  ylab("Time in Shelter (days)")

```

This graph shows that the age of an animal has little to no relevance to the amount of time said animal stays in the shelter. The regression lines show decently horizontal lines for all the pet types. The pet type with the largest correlation is dogs, with a slightly positive slope.

```{r}

#Does being old increase shelter time 

pets |> 
  mutate(AgeGroups = if_else(AgeMonths <24, "Young(<2 years old)", if_else(AgeMonths < 84, "Adult (2-7 years)", "Old(>7 years old"))) |> 
  group_by(AgeGroups) |> 
  summarise(
    median = median(TimeInShelterDays)
  )  |> 
  ggplot(aes(x = AgeGroups, y = median)) + 
  geom_col() + 
  ggtitle("Age Groups and Time in Shelter")

```

This graph displays that younger pets (under 2) tend to spend less time in the shelter. This graph was created to see whether older pets are more likely to spend a longer time in the shelter. The data does support that assumption, but not significantly, as adult pets have a similar median.

```{r}
#long stay calculate top 10% of the total time shelter
pets2 <- pets |> 
  mutate(LongStay = TimeInShelterDays >= quantile(TimeInShelterDays, 0.9, na.rm = TRUE))


#Is there a breed that typical stays longer 

LongStayBreed <- pets2 |>
  group_by(Breed) |> 
  summarise(
    n = n(),
    medianDays = median(TimeInShelterDays), longStayRate = mean(LongStay),
    avgDays  =  mean(TimeInShelterDays)
  ) |> 
  arrange(desc(medianDays))
LongStayBreed


LongStayBreed |> 
  filter(n >=30) |> 
  arrange(desc(longStayRate)) |> 
  ggplot(aes(x = longStayRate, y = reorder(Breed, longStayRate))) +
  geom_bar(stat = "identity", fill = rgb(0, 80, 136, maxColorValue = 255) ) + 
  coord_flip()+theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  xlab("Long Stay Rate") +
  ylab("Breed") + 
  ggtitle("Long Stay Rate and Breed")
```

Long Stays are the top 10% of time spent in the shelter. The graph shows that Labradors are the breed with the most Long Stay residents with around 15% of long stay pets are Labrador. The least likely to have a long stay is Poodles.

```{r}

#Does having a healthCondition impact shelter time?

pets |> 
  group_by(HealthCondition) |> 
  summarise(
    n = n(),
    avgDays = mean(TimeInShelterDays),
    medianDays = median(TimeInShelterDays)
  ) 

pets |> 
  ggplot(aes(x = factor(HealthCondition), y = TimeInShelterDays, fill = factor(HealthCondition)))+
  geom_boxplot() + 
  xlab("Health Condition") + 
  labs(fill = "Health Condition") + 
  ggtitle("Time In Shelter and Health Condition")


```

The ranges on the boxplots and the medians are relative the same thus there is no significant impact of health condition on the amount of shelter days.

```{r}

#add titles and colors 
pets |> 
  ggplot(aes(x = factor(HealthCondition), y = TimeInShelterDays, fill = factor(HealthCondition))) +
  geom_boxplot() +
  coord_flip()+
  facet_wrap(~PetType)  +
  xlab("Health Condition") + 
  ggtitle("Time In Shelter and Health Condition") + 
  labs(fill = "Health Condition")

```

The graph shows that having a health condition does not drastically impact the amount of time animal spend in the shelter regardless of the type of pet. Although the cats and dogs are slightly more likely to spend less time in the shelter if they are healthy.

3.  **Is there a breed that has a larger adoption fee? Does the adoption fee have any correlation to the size of the pet? Do pets with higher adoption fees have better health conditions/are vaccinated? Does the color of the pet impact the adoption fee?**

```{r}

# Is there a breed that has a larger adoption fee?
# Using median fee to avoid skewness from outliers
pets |> 
  group_by(Breed, PetType) |>
  summarise(
    n = n(),
    avgFee = mean(AdoptionFee),
    medianFee = median(AdoptionFee),
    .groups = 'drop'
  ) |> 
  mutate(BreedLabel = paste0(Breed, " (", PetType, ")")) |>
  arrange(desc(medianFee)) |> 
  ggplot(aes(x = reorder(BreedLabel, medianFee), y = medianFee)) + 
  geom_col(fill = "lightseagreen") + 
  coord_flip() + 
  ggtitle("Median Adoption Fee by Breed") +
  xlab("Breed (Pet Type)") +
  ylab("Median Adoption Fee ($)")

# Does the adoption fee have any correlation to the size of the pet?
cor(pets$WeightKg, pets$AdoptionFee, use = "complete.obs")
pets |> 
  ggplot(aes(x = WeightKg, y = AdoptionFee)) +
  geom_point(alpha = 0.5, color = "sienna") +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  facet_wrap(~PetType) + 
  ggtitle("Adoption Fee vs. Pet Weight by Pet Type") +
  xlab("Weight (Kg)") +
  ylab("Adoption Fee ($)")

# Do pets with higher adoption fees have better health conditions/are vaccinated?
pets |> 
  ggplot(aes(x = HealthCondition, y = AdoptionFee, fill = HealthCondition)) +
  geom_boxplot() +
  ggtitle("Adoption Fee by Health Condition") +
  xlab("Health Condition") +
  ylab("Adoption Fee ($)")

pets |> 
  ggplot(aes(x = Vaccinated, y = AdoptionFee, fill = Vaccinated)) +
  geom_boxplot() +
  scale_fill_manual(values = c("FALSE" = "tomato", "TRUE" = "springgreen4")) +
  ggtitle("Adoption Fee by Vaccination Status") +
  xlab("Vaccinated") +
  ylab("Adoption Fee ($)")

# Does the color of the pet impact the adoption fee?
pets |> 
  ggplot(aes(x = reorder(Color, AdoptionFee, FUN = median), y = AdoptionFee, fill = Color)) +
  geom_boxplot() +
  scale_fill_manual(values= c("Black" = "grey30", "Brown" = "burlywood4", 
                              "Gray" = "gray70", "Orange" = "darkorange", "White" = "wheat1")) +
  ggtitle("Adoption Fee Distribution by Color") +
  xlab("Color") +
  ylab("Adoption Fee ($)") +
  theme(legend.position = "none")

# Does adoption fee impact the likelihood of adoption?
pets |> 
  ggplot(aes(x = AdoptionLikelihood, y = AdoptionFee, fill = AdoptionLikelihood)) +
  geom_boxplot() +
  scale_fill_manual(values = c("FALSE" = "tomato", "TRUE" = "springgreen4")) +
  ggtitle("Adoption Fee by Adoption Likelihood") +
  xlab("Adoption Likelihood") +
  ylab("Adoption Fee ($)")
```

4.  **Which type of pets are most likely to be healthy (HealthCondition = 1), and how does that connect to the time spent in the shelter? Are medical consitions more common among older animals across all pet types? Are certain breeds more likely to have a medical condition?**

```{r}
# Which type of pets are most likely to be healthy (HealthCondition = FALSE)?
pets |> 
  group_by(PetType) |> 
  summarise(
    n = n(),
    HealthRate = 1 - mean(HealthCondition) 
  ) |> 
  arrange(desc(HealthRate)) |> 
  ggplot(aes(x = reorder(PetType, HealthRate), y = HealthRate)) +
  geom_col(fill = "mediumpurple3") +
  ggtitle("Proportion of Healthy Pets by Pet Type") +
  xlab("Pet Type") +
  ylab("Proportion Healthy")

# How does that connect to the time spent in the shelter?
pets |> 
  ggplot(aes(x = HealthCondition, y = TimeInShelterDays, fill = HealthCondition)) +
  geom_boxplot() +
  scale_fill_manual(values = c("FALSE" = "springgreen4", "TRUE" = "tomato")) +
  facet_wrap(~PetType) +
  ggtitle("Time in Shelter by Medical Condition and Pet Type") +
  xlab("Medical Condition") +
  ylab("Time in Shelter (Days)")

# Are medical conditions more common among older animals across all pet types?
pets |> 
  mutate(AgeGroups = if_else(AgeMonths < 24, "Young (<2 years)", 
                     if_else(AgeMonths < 84, "Adult (2-7 years)", "Old (>7 years)"))) |>
  mutate(AgeGroups = factor(AgeGroups, levels = c("Young (<2 years)", "Adult (2-7 years)", "Old (>7 years)"))) |>
  group_by(AgeGroups, PetType) |> 
  summarise(
    n = n(),
    MedicalConditionRate = mean(HealthCondition), 
    .groups = 'drop'
  ) |> 
  ggplot(aes(x = AgeGroups, y = MedicalConditionRate, fill = PetType)) +
  geom_col(position = "dodge") + 
  ggtitle("Rate of Medical Conditions by Age Group and Pet Type") +
  xlab("Age Group") +
  ylab("Proportion with Medical Condition")

# Are certain breeds more likely to have a medical condition?
pets |> 
  group_by(Breed, PetType) |>
  summarise(
    n = n(),
    MedicalConditionRate = mean(HealthCondition),
    .groups = 'drop'
  ) |> 
  mutate(BreedLabel = paste0(Breed, " (", PetType, ")")) |>
  arrange(desc(MedicalConditionRate)) |> 
  ggplot(aes(x = reorder(BreedLabel, MedicalConditionRate), y = MedicalConditionRate)) +
  geom_col(fill = "indianred") +
  coord_flip() +
  ggtitle("Proportion of Pets with Medical Conditions by Breed") +
  xlab("Breed (Pet Type)") +
  ylab("Proportion with Medical Condition")

# Does health condition impact the likelihood of adoption?
pets |> 
  group_by(HealthCondition) |> 
  summarise(
    n = n(),
    AdoptionRate = mean(AdoptionLikelihood)
  ) |> 
  ggplot(aes(x = HealthCondition, y = AdoptionRate, fill = HealthCondition)) +
  geom_col(color = "black") + 
  scale_fill_manual(values = c("FALSE" = "springgreen4", "TRUE" = "tomato")) +
  ggtitle("Adoption Rate by Health Condition") +
  xlab("Medical Condition") +
  ylab("Adoption Rate") +
  theme(legend.position = "none")
```

## Conclusion

After analyzing and cleaning the dataset, the results suggest that a pet's health condition is the most important factor that shelters can influence to improve adoption rates.

-   Health and Pet Type are The Biggest Drivers of Adoption

    -   The data show that pet type and health status are the most critical factors in adoption. Dogs and animals with no health conditions are more likely to find a home. 

-   Time in Shelter is Unpredictable

    -    Variables like a pet's age, adoption fee, and medical condition show little to no correlation with how many days they spend in the shelter.

-   Adoption Fees are Static

    -    Fees are not adjusted based on a pet's size, vaccination status, or medical needs, and the cost does not positively or negatively influence a pet's chance of being adopted.

-   Predicting Adoption Requires More

    -   Since there are few strong trends, a pet’s likelihood of being adopted is likely based on complex combinations of variables. For an accurate prediction, we would need advanced machine learning models.
